Accuracy in Parameter Estimation and Simulation Approaches for Sample Size Planning Accounting for Item Effects

### 2025-08-22 manuscript_draft.Rmd
updating disclosures
update manuscript RMD

```{r trackdown, eval = F}
# devtools::install_github("ClaudioZandonella/trackdown")
library(trackdown)
# initial set up 
# upload_file(file = "manuscript_draft.Rmd",
#       gfile = "power2stim_V2_trackdown",
#       gpath = "Studies/PSA007 SPAM-L/Manuscripts/Power",
#       shared_drive = "Psychological Science Accelerator")
# from google 
download_file(file = "manuscript_draft.Rmd", 
      gfile = "power2stim_V2_trackdown",
      gpath = "trackdown")
# to google
update_file(file = "manuscript_draft.Rmd", 
      gfile = "power2stim_V2_trackdown",
      gpath = "trackdown",
      rich_text = TRUE,
      hide_code = TRUE)
```

@@

AIPE shifts the focus away from finding a significant *p*-value to finding a parameter that is accurately measured. As discussed in Kelley and Maxwell [@kelley2003], one can estimate the sample size necessary to obtain precision in estimation of population parameters. Precision is defined as a researcher defined, sufficiently narrow, confidence interval in AIPE. For example, researchers may wish to detect a specific mean in a study, *M* = .35. They could then use AIPE to estimate the sample size needed to find a sufficiently narrow window around that mean. Therefore, they could decide that sufficiently narrow could be defined as a width of .30 or .15 on each side of the mean. They would then estimate the number of participants needed to find that level of precision. Confidence intervals are often misinterpreted [@hoekstra2014; but see @miller2016], and AIPE procedures are not designed to specify sample size for a hypothesis-driven decision (i.e., the confidence interval does not include a specific value of comparison). Instead, AIPE focuses on estimating the sample size necessary to ensure precise population parameters. Note that any particular confidence interval is not of interest within our procedure, but rather how to define the sufficiently narrow window a researcher should use for sample size estimation when designing studies with multiple items.

@@

One form of data simulation is data-driven Monte-Carlo simulation, which involves using data obtained to simulate similar datasets by drawing from the original data with replacement [@efron2000; @rousselet]. This type of simulation allows one to calculate parameter estimates, confidence intervals, and to simulate the potential population distribution, shape, and bias. Simulation is often paired with re-creating a data set with a similar structure for testing analyses and hypotheses based on proposed effect sizes or suggested population means. Generally, we would suggest starting with pilot data of a smaller sample size (e.g., 20 to 50) to understand the variability in potential items used to represent your phenomenon, especially if they are to be used in a larger study. However, given some background knowledge about the potential items, one could simulate example pilot data to use in a similar manner in our suggested procedure.

@@

One would set a minimum sample size based on our procedure steps below to ensure an appropriate minimum for precise estimates. After meeting the minimum sample size, researchers could then use sequential testing to estimate their parameter of interest after each participant's data or at regular intervals during data collection to determine whether they have achieved their expected narrow window around that parameter. A stricter criterion could be defined for a stopping rule as a maximum sample size. By defining each of these components, researchers could ensure a feasible minimum sample size, a way to stop data collection when goals have been met, and a maximum sample size rule to ensure an actual end to data collection. The maximum stopping rule could also be defined by resources (e.g., two semesters data collection), but nevertheless should be included. The advantage of sequential testing lies within research studies that use a random selection of items across participants (i.e., participants do not see all items). Across participants, data can be shifted to items with more uncertainty to increase precision of estimates for those items. Sequential testing is not a necessary component to our proposed procedure but we outline how to use the sample size estimates below to set minimum, maximum, and stopping rules. If researchers will show all items to participants, they could simply select one of the proposed estimates for their minimum required sample size. Therefore, we propose a method that leverages the ideas behind AIPE, paired with simulation and Monte Carlo simulation, to estimate the minimum and maximum proposed sample sizes and stopping rules for studies that use multiple items with expected variability in their estimates.

@@

### Calculate the Stopping Rule

@@

2)  For each item in the pilot data, calculate the standard error (SE). Select a cutoff SE that defines when items are considered accurately measured. The simulations described in the Data Simulation section will explore what criterion should be used to determine the cutoff SE from the pilot data.

@@

@@

Last, we examined the question of an appropriate SE decile. First, the 0, 1st, and 2nd deciles are likely too restrictive, providing very large estimates that do not always find a reasonable sample size in proportion to the pilot sample size, scale size, and heterogeneity. If we examine the $R^2$ values for each decile of our regression equation separately, we find that the values are all $R^2$ \> .99 with very little differences between them. Figures \@ref(fig:decile40-figure) and \@ref(fig:decile50-figure) illustrate the corrected scores for simulations at the 4th and 5th decile recommended cutoff for item standard errors. For small heterogeneity, differences in decile are minimal, while larger heterogeneity shows more correction at the 4th decile range, especially for scales with larger potential variance. Therefore, we would suggest the 4th decile to overpower each item for Step 2.

```{r decile40-figure, fig.cap = "Comparison of the cutoffs for 4th deciles across heterogeneity (columns), powering of items (rows), and scale size (color).", fig.height=8, fig.width=8}
summary_long %>% 
  filter(name == "Decile 40") %>% 
  group_by(variability, original_n, scale_size, name, power) %>% 
  summarize(new_sample = mean(new_sample)) %>% 
  ggplot(., aes(original_n, new_sample, color = scale_size)) + 
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line(aes(original_n, original_n), color = "black") + 
  geom_point() + 
  geom_line() + 
  theme_classic(base_size = 10) + 
  facet_grid(power~ variability,
             labeller = labeller(power = power.labs)) +
  ggtitle("Results for 4th deciles") +
  xlab("\nPilot Sample Size") + 
  ylab("Corrected Sample Size\n") + 
  scale_color_manual(name = "Scale Size",
                     values = c(plot_colors)) + 
  theme_bw() +
  theme(legend.position="bottom")

@@

```{r decile50-figure, fig.cap = "Comparison of the cutoffs for 5th deciles across heterogeneity (columns), powering of items (rows), and scale size (color).", fig.height=8, fig.width=8}
summary_long %>% 
  filter(name == "Decile 50") %>% 
  group_by(variability, original_n, scale_size, name, power) %>% 
  summarize(new_sample = mean(new_sample)) %>% 
  ggplot(., aes(original_n, new_sample, color = scale_size)) + 
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line(aes(original_n, original_n), color = "black") + 
  geom_point() + 
  geom_line() +
  theme_classic(base_size = 10) + 
  facet_grid(power~ variability,
             labeller = labeller(power = power.labs)) + 
  ggtitle("Results for 5th deciles") +
  xlab("\nPilot Sample Size") + 
  ylab("Corrected Sample Size\n") + 
  scale_shape_discrete(name = "Scale Size") +
  scale_color_manual(name = "Cutoff Decile",
                     values = c(plot_colors)) + 
  theme_bw() +
  theme(legend.position="bottom")

@@

@@

@@

The above simulations assume symmetric population distributions with small, medium, or large heterogeneity or a skewed distribution with the same heterogeneity. However, as shown in @pollock2018, sample rating data that appears to have a mean distribution may represent the average two separate distributions (i.e., a bimodal distribution). To examine our procedure with bimodal distributions, we estimated thirty items for three distributions: symmetric, ceiling, and floor distributions for Likert data with small heterogeneity. Bimodal data was created by selecting half of the ceiling distribution and floor distribution to combine together. The number of bimodal items was varied from 0 percent to 100 percent increasing by 10 percent increments for simulated samples. Figure \@ref(fig:bimodal-figure) shows an example of 10 of the items within a simulation that estimated that half of the Likert items would be bimodal in nature. The researcher procedure described above was then carried out using *semanticprimeR*.

@@

Second, these methods do not ensure the normal interpretation of power, focusing on finding a specific effect for a specific test, $\alpha$, and so on. As discussed in the introduction, there is not necessarily a one-to-one mapping of hypothesis to analysis; many of the estimations within a traditional power analysis are just that - best approximations for various parameters. These proposed methods and traditional power analysis could be used together to strengthen our understanding of the sample size necessary for both a hypothesis test and a well-tuned estimation. We would advise caution when oversampling (i.e., simulation from data with larger sample sizes than the pilot data) for null-hypothesis power estimates, as these may be upwardly biased in small samples creating Type-I error [@burns].

@@

- Data: All data used in this manuscript and vignettes have been cited and can be found on our repository pages (https://github.com/SemanticPriming/stimuli-power).
- Analysis Code: 
  - Manuscript repository with code and data: <https://osf.io/swmva/> or <https://github.com/SemanticPriming/stimuli-power>
  - Package repository with vignettes and data: <https://github.com/SemanticPriming/semanticprimeR>
- Pre-registration: We did not pre-register this study, as it was a simulation study.
- Materials: No materials were used.
- Funding: no funding was provided for this study. 
- Conflicts of interest: the researchers declare no conflicts of interest.

>>>>

```{r trackdown, eval = F}
# devtools::install_github("ClaudioZandonella/trackdown")
library(trackdown)
# initial set up 
source("~/GitHub/files/google_creds.R")
# upload_file(file = "manuscript_draft.Rmd",
#       gfile = "power2stim_V3_trackdown")
# from google 
download_file(file = "manuscript_draft.Rmd", 
      gfile = "power2stim_V3_trackdown",
      gpath = "trackdown")
# to google
update_file(file = "manuscript_draft.Rmd", 
      gfile = "power2stim_V3_trackdown",
      gpath = "trackdown",
      rich_text = TRUE,
      hide_code = TRUE)
```

@@

AIPE shifts the focus away from finding a significant *p*-value to finding a parameter that is accurately measured. As discussed in Kelley and Maxwell [-@kelley2003], one can estimate the sample size necessary to obtain precision in estimation of population parameters. Precision is defined as a researcher defined, sufficiently narrow, confidence interval in AIPE. For example, researchers may wish to detect a specific mean in a study, *M* = .35. They could then use AIPE to estimate the sample size needed to find a sufficiently narrow window around that mean. Therefore, they could decide that sufficiently narrow could be defined as a width of .30 or .15 on each side of the mean. They would then estimate the number of participants needed to find that level of precision.

@hoekstra2014 argued that confidence intervals are often misinterpreted [see @miller2016 for critique of this claim; see @morey2016 for original authors’ response], and AIPE procedures are not designed to specify sample size for a hypothesis-driven decision (i.e., the confidence interval does not include a specific value of comparison). Instead, AIPE focuses on estimating the sample size necessary to ensure precise population parameters. Note that any particular confidence interval is not of interest within our procedure, but rather how to define the sufficiently narrow window a researcher should use for sample size estimation when designing studies with multiple items.

@@

One form of data simulation is data-driven Monte-Carlo simulation, which involves using data obtained to simulate similar datasets by drawing from the original data with replacement [@efron2000; @rousselet]. This type of simulation allows one to calculate parameter estimates, confidence intervals, and to simulate the potential population distribution, shape, and bias. Simulation is often paired with re-creating a data set with a similar structure for testing analyses and hypotheses based on proposed effect sizes or suggested population means. Generally, we would suggest starting with pilot data of a smaller sample size (e.g., 20 to 50, see below for tests to determine the appropriate minimum) to understand the variability in potential items used to represent your phenomenon, especially if they are to be used in a larger study. However, given some background knowledge about the potential items, one could simulate example pilot data to use in a similar manner in our suggested procedure.

@@

One would set a minimum sample size based on our procedure steps below to ensure an appropriate minimum for precise estimates. After meeting the minimum sample size, researchers could then use sequential testing to estimate their parameter of interest after each participant's data or at regular intervals during data collection to determine whether they have achieved their expected narrow window around that parameter. A stricter criterion could be defined for a stopping rule (i.e., the criterion that specifies when data collection ends, such as reaching a maximum sample size or achieving a desired level of precision). By defining each of these components, researchers could ensure a feasible minimum sample size, a way to stop data collection when goals have been met, and a maximum sample size rule to ensure an actual end to data collection. The maximum stopping rule could also be defined by resources (e.g., two semesters data collection), but nevertheless should be included. The advantage of sequential testing lies within research studies that use a random selection of items across participants (i.e., participants do not see all items). Across participants, data can be shifted to items with more uncertainty to increase precision of estimates for those items. Sequential testing is not a necessary component to our proposed procedure but we outline how to use the sample size estimates below to set minimum, maximum, and stopping rules. If researchers will show all items to participants, they could simply select one of the proposed estimates for their minimum required sample size. Therefore, we propose a method that leverages the ideas behind AIPE, paired with Monte Carlo simulation, to estimate the minimum and maximum proposed sample sizes and stopping rules for studies that use multiple items with expected variability in their estimates. Sequential testing and sequential AIPE methods have a substantial literature of their own [@wald1992; @kelley2007; @kelley2018; @siegmund1985; @chow2006], and our approach is complementary rather than a replacement for those frameworks.

@@

### Define Pilot Data and Cutoff Criterion

@@

2)  For each item in the pilot data, calculate the standard error (SE). Select a cutoff SE that defines when items are considered accurately measured. The simulations described in the Data Simulation section will explore what criterion should be used to determine the cutoff SE from the pilot data. Similar concepts appear in classical estimation work where lower bounds of population standard deviations are used as benchmarks [@mukhopadhyay1980; @chattopadhyay2021].

@@

*Assumptions.* Our procedure does not assume normality of the data. To illustrate this, we simulated populations with normal, skewed, and bimodal distributions (in special considerations section). What the procedure requires is that item-level scores are sampled from a reasonably stable distribution with finite moments (i.e., mean and variance exist). The method evaluates variability in standard errors rather than relying on strict distributional forms. Thus, while we show results under both normal and non-normal populations, the logic of the procedure is distribution-agnostic provided these basic moment conditions are met.

@@

Last, we examined the question of an appropriate SE decile. The minimum and first two deciles are likely too restrictive, providing very large estimates that do not always find a reasonable sample size in proportion to the pilot sample size, scale size, and heterogeneity. If we examine the $R^2$ values for each decile of our regression equation separately, we find that the values are all $R^2$ \> .99 with very little differences between them. Figures \@ref(fig:decile40-figure) and \@ref(fig:decile50-figure) illustrate the corrected scores for simulations at the 4th and 5th decile recommended cutoff for item standard errors. For small heterogeneity, differences in decile are minimal, while larger heterogeneity shows more correction at the 4th decile range, especially for scales with larger potential variance. Therefore, we would suggest the 4th decile to overpower each item for Step 2.

```{r decile40-figure, fig.cap = "Comparison of the cutoffs for 4th deciles across heterogeneity (columns), powering of items (rows), and scale size (color).", fig.height=8, fig.width=8}
summary_long %>% 
  filter(name == "Decile 40") %>% 
  group_by(variability, original_n, scale_size, name, power) %>% 
  summarize(new_sample = mean(new_sample)) %>% 
  ggplot(., aes(original_n, new_sample, color = scale_size)) + 
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line(aes(original_n, original_n), color = "black") + 
  geom_point() + 
  geom_line() + 
  theme_linedraw(base_size = 10) + 
  facet_grid(power~ variability,
             labeller = labeller(power = power.labs)) +
  ggtitle("Results for 4th deciles") +
  xlab("\nPilot Sample Size") + 
  ylab("Corrected Sample Size\n") + 
  scale_color_manual(name = "Scale Size",
                     values = c(plot_colors)) + 
  theme_bw() +
  theme(legend.position="bottom")

@@

```{r decile50-figure, fig.cap = "Comparison of the cutoffs for 5th deciles across heterogeneity (columns), powering of items (rows), and scale size (color).", fig.height=8, fig.width=8}
summary_long %>% 
  filter(name == "Decile 50") %>% 
  group_by(variability, original_n, scale_size, name, power) %>% 
  summarize(new_sample = mean(new_sample)) %>% 
  ggplot(., aes(original_n, new_sample, color = scale_size)) + 
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line(aes(original_n, original_n), color = "black") + 
  geom_point() + 
  geom_line() +
  theme_linedraw(base_size = 10) + 
  facet_grid(power~ variability,
             labeller = labeller(power = power.labs)) + 
  ggtitle("Results for 5th deciles") +
  xlab("\nPilot Sample Size") + 
  ylab("Corrected Sample Size\n") + 
  scale_shape_discrete(name = "Scale Size") +
  scale_color_manual(name = "Cutoff Decile",
                     values = c(plot_colors)) + 
  theme_bw() +
  theme(legend.position="bottom")

@@

These corrected values represent the final recommended minimum sample sizes for achieving the specified precision thresholds (e.g., 90%, 95%). Researchers can use these numbers directly to determine the target N for their main study, and to define a feasible range of sample sizes. In practice, the corrected N can serve as the minimum sample size, while researchers may also set a maximum sample size based on available resources, and optionally apply sequential monitoring rules to stop data collection early if the desired precision is reached. These corrected estimates translate the pilot-based procedure into concrete targets that researchers can use when planning and monitoring data collection.

@@

### Pilot Sample Size

```{r pilot-sim-data}
ps_DF <- import("simulation_study/simulation_pilot/pilot_simulation_final.csv") %>% 
  select(-source_file) %>% 
  group_by(pilot_size, target_percent, scale_size, variance_size) %>% 
summarise(
    avg_final_N = mean(final_N, na.rm = TRUE),
    sd_final_N = sd(final_N,   na.rm = TRUE),
    se_final_N = sd(final_N,   na.rm = TRUE)/sqrt(sum(!is.na(final_N))),
    avg_pct_below_pop = mean(pct_items_below_pop, na.rm = TRUE),
    se_pct_below_pop = sd(pct_items_below_pop, na.rm = TRUE)/sqrt(sum(!is.na(pct_items_below_pop))),
    avg_pct_below_pilot = mean(pct_items_below_pilot, na.rm = TRUE),
    se_pct_below_pilot = sd(pct_items_below_pilot, na.rm = TRUE)/sqrt(sum(!is.na(pct_items_below_pilot))),
    avg_delta_pct_pop = mean(delta_pct_pop, na.rm = TRUE),
    avg_se_bias = mean(se_bias, na.rm = TRUE),
    avg_se_mae = mean(se_mae, na.rm = TRUE),
    avg_se_rmse = mean(se_rmse, na.rm = TRUE),
    avg_se_mean_rel_err = mean(se_mean_rel_err, na.rm = TRUE),
    avg_cov_within_5pct = mean(se_cov_within_5pct, na.rm = TRUE),
    avg_cov_within_10pct = mean(se_cov_within_10pct, na.rm = TRUE),
    avg_cutoff_error = mean(cutoff_error, na.rm = TRUE),
    avg_cutoff_rel_error = mean(cutoff_rel_error, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  mutate(scale_size = factor(scale_size, 
                             levels = c("small", "medium", "large"),
                             labels = c("Likert", "Percent", "Milliseconds")),
         variance_size = factor(variance_size, 
                             levels = c("small", "medium", "large"),
                             labels = c("Small Heterogeneity", "Medium Heterogeneity", "Large Heterogeneity")))
```

Smaller pilot samples may be expected to show greater variability than larger samples or the full population. This variability could influence the sample size recommendations generated by our procedure. At the same time, the optimal size of the pilot sample was an open question in our framework. To address these questions, we ran simulations designed to identify the smallest pilot sample size that would minimize variability across researchers while also reducing bias when comparing pilot-derived estimates to population values. In these simulations, 100 researchers each drew a pilot sample from the same target population and applied our procedure. We generated nine populations (the combinations of scale size and heterogeneity described above), each with 30 items (`simulate_population`). For each population, we varied the pilot sample size across 20, 25, 30, 35, and 40. Each simulated researcher then ran 500 replications, drawing between 20 and 300 participants in increments of 5 (`simulate_samples`).

We evaluated three statistics from these simulations:

1.  Variability of the recommended sample size. For each target level, we calculated the average recommended *N* and the standard deviation of that recommendation. Large standard deviations indicate instability, meaning different researchers could receive very different recommendations from similar pilot studies.
2.  Bias in item standard errors. For each item, we compared the estimated *SE* from the sample to the true *SE* from the population. Smaller bias reflects closer agreement between pilot-based estimates and population values. Because larger sample sizes and heterogeneity simulations will naturally show larger SEs in general, we calculated relative error by subtracting estimated *SE* minus the true *SE* divided by the true *SE*.
3.  Cut-off bias. The cut-off was fixed at the 4th decile in the population. For each pilot, we calculated the difference between the pilot-derived cut-off and the population cut-off. Smaller values indicate that pilot-derived thresholds align more closely with the population criterion. Again, we used the relative error by dividing by the true *SE* cut off score to be able to compare across simulations.

Results consistently indicated that smaller pilots produced less stable and a bit more biased recommendations. As shown in Figure \@ref(fig:fig-pilot-sd-n), the standard deviation of the recommended sample size was higher at pilot sizes of 20 and 25, particularly in large-scale, high-variance populations. This instability means that two researchers running similar small pilots could easily obtain divergent recommended *N*s. By contrast, pilot sizes of 30 or larger substantially reduced this variability, and further increases to 35 or 40 provided little additional gain. Bias measures told a similar story. Figure \@ref(fig:fig-pilot-bias) shows that average item *SE* bias was near zero across most conditions, but small pilots (20–25) tended to produce more negative bias in large-scale populations. At pilot sizes of 30 and above, *SE* bias converged toward zero across all conditions. Likewise, Figure \@ref(fig:fig-pilot-cutoff) shows that relative cut-off bias was largest when pilot sizes were small, again especially under high-variance conditions, but shrank considerably by pilot size 30 and remained stable thereafter. Accordingly, we propose \~30 participants as a practical lower bound for pilot studies using our procedure: large enough to ensure stability, accuracy, and calibration across a wide range of scale and variance conditions, but small enough to remain efficient.

```{r fig-pilot-sd-n, fig.cap = "Average standard deviation of suggested sample sizes for small, medium, and large scale sizes and heterogeneity.", fig.height=6, fig.width=8, warning = F, message = F}
ggplot(ps_DF %>% 
         filter(target_percent == 80), aes(x = pilot_size, y = sd_final_N,
                  color = variance_size, group = variance_size)) +
  geom_point() +
  #geom_line() +
  # geom_errorbar(aes(ymin = avg_final_N - se_final_N,
  #                   ymax = avg_final_N + se_final_N), width = 0.2) +
  facet_wrap(~ scale_size) +
  labs(x = "Pilot Sample Size", y = "Average SD of N") +
  theme_linedraw() + 
  theme_bw() + 
  theme(legend.position = "bottom") + 
  scale_color_manual(name = NULL, values = c(plot_colors)) 
```

```{r fig-pilot-bias, fig.cap = "Average relative standard error bias calculated for each simulation of size and heterogeneity combination. Relative error subtracts the pilot sample size SE for each item by the true SE for each item and then divides by the true SE for each item to normalize across scales. ", fig.height=6, fig.width=8, warning = F, message = F}
ggplot(ps_DF %>% 
         filter(target_percent == 80), aes(x = pilot_size, y = avg_se_mean_rel_err,
                  color = variance_size, group = variance_size)) +
  geom_point() +
  #geom_line() +
  # geom_errorbar(aes(ymin = avg_final_N - se_final_N,
  #                   ymax = avg_final_N + se_final_N), width = 0.2) +
  facet_wrap(~ scale_size) +
  labs(x = "Pilot Sample Size", y = "Average Relative SE Bias") +
  theme_linedraw() + 
  theme_bw() + 
  theme(legend.position = "bottom") + 
  scale_color_manual(name = NULL, values = c(plot_colors)) 
```

```{r fig-pilot-cutoff, fig.cap = "Average relative cut-off score bias for scale size and heterogeneity combinations. Relative cut-off scores indicate the true population 4th decile SE compared to the pilot sample SE cutoff at the 4th decile, divided by the population SE. ", fig.height=6, fig.width=8, warning = F, message = F}
ggplot(ps_DF %>% 
         filter(target_percent == 80), aes(x = pilot_size, y = avg_cutoff_rel_error,
                  color = variance_size, group = variance_size)) +
  geom_point() +
  #geom_line() +
  # geom_errorbar(aes(ymin = avg_final_N - se_final_N,
  #                   ymax = avg_final_N + se_final_N), width = 0.2) +
  facet_wrap(~ scale_size) +
  labs(x = "Pilot Sample Size", y = "Average Relative Cut-Off Bias") +
  theme_linedraw() + 
  theme_bw() + 
  theme(legend.position = "bottom") + 
  scale_color_manual(name = NULL, values = c(plot_colors)) 
```

@@

The above simulations assume symmetric population distributions with small, medium, or large heterogeneity or a skewed distribution with the same heterogeneity. However, as shown in @pollock2018, rating data that appears normally distributed at the aggregate level may in fact reflect the average of two distinct underlying distributions (i.e., a bimodal distribution). In such cases, a pilot sample may misleadingly suggest a single “middle” value that does not represent either true subgroup. To examine our procedure with bimodal distributions, we estimated thirty items for three distributions: symmetric, ceiling, and floor distributions for Likert data with small heterogeneity. Bimodal data was created by selecting half of the ceiling distribution and floor distribution to combine together. The number of bimodal items was varied from 0 percent to 100 percent increasing by 10 percent increments for simulated samples. Figure \@ref(fig:bimodal-figure) shows an example of 10 of the items within a simulation that estimated that half of the Likert items would be bimodal in nature. The researcher procedure described above was then carried out using *semanticprimeR*.

@@

Second, these methods do not ensure the normal interpretation of power, focusing on finding a specific effect for a specific test, $\alpha$, and so on. As discussed in the introduction, there is not necessarily a one-to-one mapping of hypothesis to analysis; many of the estimations within a traditional power analysis are just that - best approximations for various parameters. These proposed methods and traditional power analysis could be used together to strengthen our understanding of the sample size necessary for both a hypothesis test and a well-tuned estimation. We would advise caution when resampling pilot data (e.g., oversampling beyond the available pilot sample) for power estimation. As @burns note, resampling introduces estimation bias that increases as simulated sample sizes approach the full pilot size. This bias can occur not only for hypothesis tests but also when estimating descriptive statistics such as means, and it may distort power estimates, particularly in small samples.

@@

-   Funding: no funding was provided for this study.
-   Conflicts of interest: the researchers declare no conflicts of interest.
-   Ethics approval: not applicable, simulation study.
-   Consent to participate: not applicable, simulation study.
-   Consent for publication: all authors approved the manuscript, no participant approval necessary.
-   Data: All data used in this manuscript and vignettes have been cited and can be found on our repository pages (<https://github.com/SemanticPriming/stimuli-power>).
-   Materials: No materials were used.
-   Analysis Code:
    -   Manuscript repository with code and data: <https://osf.io/swmva/> or <https://github.com/SemanticPriming/stimuli-power>
    -   Package repository with vignettes and data: <https://github.com/SemanticPriming/semanticprimeR>
-   Pre-registration: We did not pre-register this study, as it was a simulation study.
-   Author contributions are printed at the top of the manuscript automatically in *papaja*.

### 2025-08-22 manuscript_draft.tex
update manuscript RMD

The above simulations assume symmetric population distributions with small, medium, or large heterogeneity or a skewed distribution with the same heterogeneity. However, as shown in @pollock2018, rating data that appears normally distributed at the aggregate level may in fact reflect the average of two distinct underlying distributions (i.e., a bimodal distribution). In such cases, a pilot sample may misleadingly suggest a single ``middle'' value that does not represent either true subgroup. To examine our procedure with bimodal distributions, we estimated thirty items for three distributions: symmetric, ceiling, and floor distributions for Likert data with small heterogeneity. Bimodal data was created by selecting half of the ceiling distribution and floor distribution to combine together. The number of bimodal items was varied from 0 percent to 100 percent increasing by 10 percent increments for simulated samples. Figure \ref{fig:bimodal-figure} shows an example of 10 of the items within a simulation that estimated that half of the Likert items would be bimodal in nature. The researcher procedure described above was then carried out using \emph{semanticprimeR}.

@@

>>>>

The above simulations assume symmetric population distributions with small, medium, or large heterogeneity or a skewed distribution with the same heterogeneity. However, as shown in Pollock (2018), rating data that appears normally distributed at the aggregate level may in fact reflect the average of two distinct underlying distributions (i.e., a bimodal distribution). In such cases, a pilot sample may misleadingly suggest a single ``middle'' value that does not represent either true subgroup. To examine our procedure with bimodal distributions, we estimated thirty items for three distributions: symmetric, ceiling, and floor distributions for Likert data with small heterogeneity. Bimodal data was created by selecting half of the ceiling distribution and floor distribution to combine together. The number of bimodal items was varied from 0 percent to 100 percent increasing by 10 percent increments for simulated samples. Figure \ref{fig:bimodal-figure} shows an example of 10 of the items within a simulation that estimated that half of the Likert items would be bimodal in nature. The researcher procedure described above was then carried out using \emph{semanticprimeR}.

@@

\bibitem[\citeproctext]{ref-pollock2018}
Pollock, L. (2018). Statistical and methodological problems with concreteness and other semantic variables: A list memory experiment case study. \emph{Behavior Research Methods}, \emph{50}(3), 1198--1216. \url{https://doi.org/10.3758/s13428-017-0938-y}
